---
title: "Assignment"
author: "Aidan Snyder"
date: "2023-02-16"
output: html_document
---

```{r message=FALSE}
library(tidyverse)
```

```{r}
getwd()
```

## 1) Load and Check Data

```{r message=FALSE}
gapminder <- read_delim("gapminder.csv.bz2")
```

```{r}
setwd("info201/ps5-rmarkdown")
```


```{r}
view(gapminder)
```

### Load data. How many rows/columns do we have?

```{r}
nrow(gapminder)
ncol(gapminder)
```

### Print a small sample of data. Does it look OK?

```{r message=FALSE}
gapminder %>% 
  sample_n(10)

```

## 2) Descriptive Statistics


### How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r}
length(unique(gapminder$name))

```

```{r}
length(unique(gapminder$iso2))

```


```{r}
length(unique(gapminder$iso3))
```

### If you did this correctly, you saw that there are more names than 
### iso-2 codes, and there are even more iso3 -codes. What is going on?
### Can you find it out?

### A.) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?

```{r}

gapminder %>% 
  group_by(iso2) %>% 
  summarise(names_for_iso2 = n_distinct(name)) %>% 
  select(iso2, names_for_iso2) %>% 
  arrange(desc(names_for_iso2))

gapminder %>% 
  filter(is.na(iso2)) %>% 
  filter(!is.na(name)) %>% 
  select(name)


```

Namibia

### (B) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?Hint: two of these entitites are CHANISL and NLD CURACAO.


```{r}


gapminder %>% 
  filter(is.na(name)) %>% 
  group_by(name) %>% 
  filter(is.na(iso2)) %>% 
  distinct(iso3)
```
CHANISL, GBM, KOS, NLD_CURACAO

### What is the minimum and maximum year in these data?

```{r}

min(gapminder$time, na.rm = TRUE)

max(gapminder$time, na.rm = TRUE)

```


## 3 CO2 emissions


### 1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?

```{r}
gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2)) %>% 
  summarise(NA_per_year = sum(is.na(co2))) %>% 
  arrange(desc(NA_per_year))

gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2_PC)) %>% 
  summarise(NAPC_per_year = sum(is.na(co2_PC))) %>% 
  arrange(NAPC_per_year)


```
The most missing includes 2017,2018,2019

The least includes 2012, 2013, and 2014

### 2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.

```{r}
gapminder %>% 
  filter(iso3 ==c( "USA", "CHN", "IND",  "ARG", "BEL")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2, col=iso3))+
  geom_point(aes(x=time, y = co2, col=iso3))



  
```
I see that every country has increased relatively over time, but China has skyrocketted in the last 20 years

###3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?

```{r}

gapminder %>% 
  filter(iso3 ==c( "USA", "CHN", "IND",  "ARG", "BEL")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2_PC, col=iso3))+
  geom_point(aes(x=time, y = co2_PC, col=iso3))

```
This suggests that per capita, most countries are on the decline or increasing very slowly, with the exception of China

###4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.Note: just compute averages over countries and ignore the fact that countries are of different size.Hint: Americas 2016 should be 4.80.

```{r}
gapminder %>% 
  group_by(region, time) %>%
  filter(!is.na(co2_PC)) %>% 
  summarize(avg_region_co2pc = mean(co2_PC))
 

```

### 5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.

```{r}


avg_region_co2pc <- gapminder %>% 
  group_by(region, time)%>%
  filter(!is.na(co2_PC)) %>% 
  summarize(avg_region_co2pc = mean(co2_PC)) 

avg_region_co2pc %>% 
  filter(!is.na(region)) %>% 
  filter(time %in% c(1960, 2016)) %>% 
  ggplot(aes(x = region, y = avg_region_co2pc, fill=time)) +
  geom_col(position = "dodge")
  
  
  


```


### 6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).

```{r}

gapminder_2016 <- gapminder %>% 
  filter(time == 2016)

largest2016 <- gapminder_2016 %>%
  group_by(region) %>%
  arrange(desc(co2_PC)) %>%
  head(3)

print(largest2016)

smallest2016 <- gapminder_2016 %>% 
   group_by(region) %>%
  filter(!is.na(co2_PC)) %>% 
  arrange(desc(co2_PC)) %>%
  tail(3)

print(smallest2016)

```

The largest include Qatar, Netherlands/Curacao, and Trinidad and Tobago

The smallest are Burundi, Somalia, and the Democratic Republic of the Congo

## 4 GDP per capita 

### 1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. Comment what do you see there.


```{r}
 
gapminder_1960 <- gapminder %>% 
  filter(time == 1960) %>% 
  filter(!is.na(region))


ggplot(gapminder_1960)+
  geom_point(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region))

```
As gross domestic product increases, life expectancy is also increasing 

### 2. (4pt) Make a similar plot, but this time use 2019 data only.

```{r}

gapminder_2019 <- gapminder %>% 
  filter(time == 2019) %>% 
  filter(!is.na(region))


ggplot(gapminder_2019)+
  geom_point(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region))

```

### 3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?


Total population and GDP per capita for countries have both increased. 
Life expectancy across most countries has also increased

### 4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? Note: here as average I mean just average over countries, ignore the fact that countries are of different size.


```{r}

gapminder_1960 %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  group_by(region) %>% 
  summarize(le_1960_avg = mean(lifeExpectancy))

gapminder_2019 %>% 
   filter(!is.na(lifeExpectancy)) %>% 
  group_by(region) %>% 
  summarize(le_2019_avg = mean(lifeExpectancy))

  

```

Yes, the results fit with what I see on the figures

###5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see. Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up!


```{r}
gapminder %>% 
  filter(!is.na(region)) %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  group_by(region, time) %>% 
   filter(time == 1960 | time == 2019) %>% 
  mutate(avg_le = mean(lifeExpectancy)) %>% 
  group_by(region) %>% 
  mutate(le_1960 = lag(avg_le, order_by = time)) %>% 
  summarise(avg_le_growth = avg_le - le_1960) %>% 
  filter(!avg_le_growth == 0) %>% 
  arrange(desc(avg_le_growth))
  
  
  
  




```

Asia and Africa had the most life expectancy growth, with the Americas and Oceania following behind, with Europe in last. One can argue that Europe grew the least as they were more developed in 1960 than other Continents. 

### 6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!

```{r}

gapminder %>% 
  filter(time == 1960 | time == 2019) %>% 
  group_by(time) %>% 
  ggplot(aes(GDP_PC, fill = factor(time))) +
  geom_histogram()
```


### 7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.) Hint: check out the function rank()! Hint2: 17 for 1960.

```{r}



gapminder %>% 
  select(name, time, lifeExpectancy) %>% 
  filter(name != "", !is.na(time), !is.na(lifeExpectancy)) %>% 
  filter(time == "1960") %>% 
  mutate(rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America")


gapminder %>% 
  select(name, time, lifeExpectancy) %>% 
  filter(name != "", !is.na(time), !is.na(lifeExpectancy)) %>% 
  filter(time == "2019") %>% 
  mutate(rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America")





```

### 8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? Hint: 0.0904 for 1960.

```{r}
gapminder_1960 %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(!is.na(name))


```
17/188 = 0.0904

```{r}
gapminder_2019 %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(!is.na(name)) 

  
```
47/196 = 0.2398

### Finally tell us how many hours did you spend on this PS.

20+ hours

